The role of cognateness in non-native spoken word recognition

### 2021-12-16 Rmd/manuscript.Rmd
Fix error when joining stimuli info and participant responses

author:
- name: Gonzalo Garcia-Castro
  affiliation: '1'
  corresponding: yes
  address: Ramon Trial Fargas, 25-27, 08005 Barcelona
  email: gonzalo.garciadecastro@upf.edu
  role: 
    - Conceptualization
    - Data analysis
    - Writing - Original Draft Preparation
    - Writing - Review & Editing
- name: Serene Siow
  affiliation: '2'
  role:
    - Conceptualization
    - Writing - Review & Editing
- name: Nuria Sebastian-Galles
  affiliation: '1'
  role:
    - Conceptualization
    - Writing - Review & Editing
- name: Kim Plunkett
  affiliation: '2'
  role:
    - Conceptualization
    - Writing - Review & Editing
affiliation:
  - id            : "1"
    institution   : "Center for Brain and Cognition, Universitat Pompeu Fabra"
  - id            : "2"
    institution   : "Department of Experimental Psychology, University of Oxford"

authornote: |
  GG and SS contributed equally and share first authorship.

abstract: |
  There is compelling evidence that bilinguals access their lexicon in a language non-selective way. For example, bilinguals produce and translate cognates (words whose translation in the other language is form-similar) faster than non-cognates, suggesting that the phonology of both languages interact during word production and comprehension (Costa et al., 2000; Christoffels et al., 2006). Previous literature on this effect has often relied on measures of overall form-similarity to categorise words into cognates andnon-cognates, such as the Levenshtein distance. These measures partially ignore potential sources of variability during lexical access like vowel vs. consonant overlap, overlap in stressed syllables, onset overlap or the distance in features between both translations. In this study, we explored the impact of some of these variables on non-native word translation task: Spanish and English participants listened to non-native words (Catalan or Spanish) and were prompted to type their translation in their native language. Critically, participants where unfamiliar with the testing language, ensuring that they were not able to translate words based on previous knowledge on their meaning, leaving phonological information as the only cue participants were able to exploit to translate words to their native language correctly. We analysed the probability of correct translations, adjusting for the amount of vowel overlap, consonant overlap, overlap at stress position, overlap at onset, and distance in features between replaced phonemes, including the average frequency of phonological neighbours of the target words are a covariate.
  <!-- https://tinyurl.com/ybremelq -->

keywords: cognate, word recognition, translation, non-native, spoken word recognition
wordcount: X

bibliography:
  - r-references.bib
  - references.bib
floatsintext: yes
figurelist: no
tablelist: no
footnotelist: no
linenumbers: yes
mask: no
draft: no
csl: "apa7.csl"
documentclass: apa7
classoption: man
output: papaja::apa6_docx
---

@@

set.seed(42)
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  results = "asis", 
  dpi = 500
)

@@

Even for highly proficient bilinguals, listening to non-native speech is more costly than listening to native speech [@lecumberri2010non]. One of the sources of this increased difficulty stems from the potential mismatch between the phonology of the native and the non-native languages: some acoustic features embedded in the non-native speech signal may not overlap perfectly with any phonemic category in the listener's native language. For example, a Spanish native listening to French may encounter the word /pɔʁt/ (*porte*), translation of *door* in French. The voiced uvular fricative consonant /ʁ/ and the open-mid back rounded vowel /ɔ/ do not exist in Spanish. This mismatch at the phonological level increases the uncertainty under which lexical selection takes place during word recognition. Yet, non-native listeners--even those with low proficiency--are rarely completely naïve to the language they are listening to. All languages share, to some extent, similarities at the lexical level, frequently due to their typological closeness, and these similarities can be exploited by non-native listeners to process the speech signal. In the present research we investigate the extent to which adult listeners rely on this similarity during lexical access, and we do so capitalising on the role of *cognates*.

Cognates are cross-language synonyms whose form (e.g., phonology, orthography, sign) is similar. The source of such similarity is frequently attributed to a shared etymological origin. Romance languages such as Spanish and Catalan share many cognates [@schepens2012distributions], as in the case of *puerta* and *porta* (*door* in Spanish and Catalan, respectively)^[Some form-similar cross-language synonyms are technically not cognates. For example, *sun* and *sol* (in Spanish), share their phonological onset, but their etymology points to different origins. We will use the word *cognateness* to include all form-similar cross-language synonyms for simplicity. It is highly implausible that etymology plays a direct role on language perception if it is not via form-similarity, since it is not necessary for participants in psycholinguistic experiments to be aware of the etymology of the words they encounter in the tasks to be subject to the effect of form-similarity.]. Cognates play a pivotal role in virtually all models of bilingual lexical processing because they provide evidence that bilinguals access the lexicon in a language non-selective way. For example, pictures are named faster by Spanish-Catalan bilinguals when their associated labels in both languages are cognates (e.g., *puerta*-*porta*), compared to when their labels are non-cognates (*mesa*-*taula*, *table* in Spanish and Catalan) [@costa2000cognate]. This cognate advantage is not restricted to production, but rather extends to word recognition [@thierry2007brain; @midgley2011effects], word learning [@de2000hard; @lotto1998effects; @valente2018does], and word translation [@christoffels2006memory].

Early theoretical accounts of this phenomenon pointed that the impact of cognateness on lexical processing is dependent on bilinguals' proficiency in their second language (L2) [@potter1984lexical; see @chen1989patterns for converging evidence]. As learners become more proficient, the connection between L2 representations and their meaning grows stronger, and L2 word processing becomes less reliant on the mediation of L1 representations. The Revised Hierarchical Model [RHM, @kroll1994category] captured this hypothesis and predicted that translating words from L1 to L2 (*forward* translation) should take longer than translating from L2 to L1 (*backward* translation). The rationale behind this prediction is that backward translation relies more strongly on direct word-word links between L1 and L2 representations, whereas forward translation relies more strongly on the mediation between the concept and the two word forms. One of the consequences of this prediction is that backward translation should be more sensitive to the form similarity between the L1 and the L2 representations: cognate words should be retrieved faster than non-cognate words during backward translation.

To test these predictions, @degroot1994forward and @de1992determinants asked Dutch natives with high (yet non-native) English proficiency to translate words from either Dutch to English (forward translation, L1 to L2) or from English to Dutch (backward translation, L2 to L1). Participants were presented with written words in English or Dutch, and asked to speak out loud their translation in the other language as fast as possible. Results showed that translation times and accuracy were roughly equivalent  in forward and backward translation, and that semantic variables (e.g., concreteness) were positively associated with participants' performance, especially in the forward translation condition. Cognates were translated equally fast in both conditions, but non-cognates were translated faster during backward than forward translation, and translation of cognates was less affected  by semantic variables than non-cognates. These findings suggest that both conceptually-mediated and direct translation routes are active during translation, although backward translation relies more strongly on direct links between L1 and L2 representations, which makes it more sensitive to cognateness (than forward translation). Interestingly, higher-proficiency bilinguals tested in the same task showed similar results. However, subsequent studies did not find differences in participant's performance during forward and backward translation, or even found better performances in forward translation [@christoffels2006memory; @christoffels2013language], contrary to the predictions of the RHM. Further models built upon this model, introducing a wider range of connections in their theorised lexica.

Since the RHM model was proposed, later studies brought attention to the role of the network of connections that words establish with each other at the form-level (phonological or orthographic) or at the conceptual level. For instance, the Neighbourhood Activation Model [NAM, @luce1998recognizing] highlighted the fact that lexical selection is affected by the number of the phonological neighbours of the selected word. Words surrounded by a larger number of phonological neighbours (only differing in one phoneme from the target words) are responded more slowly and less accurately than words from sparser phonological neighbourhoods [@goldinger1989priming; @luce1990similarity], especially when such neighbours; lexical frequency is higher [@luce1998recognizing]. This finding poses an important question in the field of bilingualism research: do word representations in one language form part of the phonological or orthographic neighbours of the other language? The Bilingual Interactive Activation [BIA+, @van1998orthographic] proposes that lexical representations establish both excitatory and inhibitory connections with other representations from the same or the other language, forming an integrated lexicon for both languages [@mcclelland1981interactive]. Evidence supporting such integrated lexicon is provided by the fact that bilinguals' performance in word recognition tasks is sensitive to the orthographic neighbourhood density of the translation of the presented word. @van1998orthographic presented bilingual participants with words in only one language (target language) in a lexical decision task. When the translation of the presented word in the non-target language was part of a large neighbourhood, reaction times were slower. This provides evidence that the presented words activated orthographic neighbourhoods in the non-target language, which slowed down word recognition.

A more recent model of bilingual lexical processing is Multilink @dijkstra2019multilink, which integrates and formalises previous claims and predictions on how an interactive account of bilingual lexical access impacts word recognition, production, and more relevant to the aims of the present study, translation. This model assumes an integrated lexicon in which word representations from the two languages establish connections in a way that is similar to what words of a single language do. Such model also assumes that during backward translation, the presented word in L2 can activate form-similar words in L1 that compete with the target word in L1 and it proposes that translation equivalents are exclusively connected through their shared concept, contrary to the RHM model (which suggests that translation equivalents are connected by excitatory connections at the lexical level; e.g., translation can be done via word-word connections). In addition, this model suggests that the strength of the association between L2 representations and their concept is a function of language users' proficiency in L2. A remarkable correlation between Multilink simulations of @christoffels2006memory translation elicitation task and behavioural data with Dutch-English bilinguals has been reported, showing forward translation being faster than backward translation, contrary to the predictions made by the RHM model. Importantly, the model also generates data supporting the claim that cognates are translated faster than non-cognates, in line with previous literature on the cognate advantage during translation. Overall, results from Multilink from this model are compatible with an integrated bilingual lexicon and with the existence of a facilitatory effect of cross-language similarity during translation.

The fact that Multilink fitted the experimental data from @christoffels2006memory with remarkable success, even in absence of word-word connections between translation equivalents might put into question the necessity of such connections during translation, in favour of a simpler model in which backward translation relies exclusively on the association between L2 and L1 word forms via their shared concept. This claim, however, should be taken. It should be taken into account, however, that @christoffels2006memory only tested fairly balanced bilinguals (i.e., bilinguals with high proficiency in L2). Additionally, data generated from Multilink model assumed similar vocabulary sizes across languages and, perhaps more critically in this regard, that word-concept connections were equally strong for L1 and L2 representations. These conclusions should therefore be taken with caution when extended to unbalanced bilinguals. It is low-proficiency bilinguals whom the RHM model predicts strong reliance on word-word connections during translation, especially during backward translation. This prediction remains untested my Multilink.

The role of this study was to investigate the plausibility of the lexical route as a mechanism exploited by monolingual adults during backward translation. Monolinguals can be considered a particular (extreme) case of unbalanced bilingualism, in which their vocabulary size in L2 is null, and word-concept connections are absent. In this case, monolinguals can only use the lexical route to succeed in the translation task, and they can only do so by exploiting the cognateness of the words presented and their translation in the native language. If participants are able to translate words in L2, even with no knowledge of the language it belongs to, based on their form-similarity with the target word, this would suggest that the lexical route is in place for low-proficiency bilinguals. If participants are not able to translation words from L2 to L1 regardless of their form-similarity, this would suggest that low-proficiency bilinguals rely entirely on word-concept connections for backward translation.

@@

@@

```{r stimulilengths, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
lengths <- stimuli %>% 
  select(trial_id, group, ipa1, ipa2, word1, word2) %>% 
  mutate_at(vars(ipa1:word2), nchar) %>% 
  group_by(group) %>% 
  summarise_at(vars(ipa1, ipa2, word1, word2), list(mean = mean, sd = sd, min = min, max = max)) %>%
  split(.$group) %>% 
  map(select, -group) %>% 
  set_names(make_clean_names(names(.))) %>% 
  map(unlist)
```

@@

We retrieved PTHN scores from the CLEARPOND database [@marian2012clearpond]. PTHN scores indicate the number of phonological neighbours of the target word with higher lexical frequency, as indicated by its score in the corresponding SUBTLEX database. CLEARPOND defines a phonological neighbour as a word whose phonological transcription in International Phonetic Alphabet (IPA) format (generated from eSPEAK, [http://espeak.sourceforge.net/](http://espeak.sourceforge.net/)) differs from that of the target word in only one addition, deletion, or substitution. PTHN scores in CLEARPOND measure have been calculated using corpora of similar size across language, allowing reliable cross-language comparisons.

@@

Participants listened to one audio file in each trial. This audio file corresponded to a word in Catalan (for Spanish speakers, and for English participants allocated in the Catalan condition) or Spanish (for English speakers allocated in the Spanish condition). The audio files were the same ones used in child experiments conducted in the Laboratori de Recerca en Infància of Universitat Pompeu Fabra (Barcelona, Spain). These audio files were recorded by a proficient Catalan-Spanish female bilingual from the Metropolitan Area of Barcelona in a child-directed manner. Catalan and Spanish words were recorded at 44,100 Hz in separate files in the same session, and then de-noised using Audacity and normalised at peak intensity using Praat [@broersma2021praat]. The average duration of the audios was `r printnum(mean(stimuli$duration, na.rm = TRUE))` (*SD* = `r printnum(sd(stimuli$duration, na.rm = TRUE))`, Range = `r printnum(min(stimuli$duration, na.rm = TRUE))`-`r printnum(max(stimuli$duration, na.rm = TRUE))`). The average duration of the Catalan audios was `r printnum(mean(stimuli$duration[stimuli$group %in% c("ENG-CAT")], na.rm = TRUE))` seconds (*SD* = `r printnum(sd(stimuli$duration[stimuli$group %in% c("ENG-CAT")], na.rm = TRUE))`, Range = `r printnum(min(stimuli$duration[stimuli$group %in% c("ENG-CAT")], na.rm = TRUE))`-`r printnum(max(stimuli$duration[stimuli$group %in% c("ENG-CAT")], na.rm = TRUE))`), and the average duration of the Spanish audios was `r printnum(mean(stimuli$duration[stimuli$group %in% c("ENG-SPA")], na.rm = TRUE))` seconds (*SD* = `r printnum(sd(stimuli$duration[stimuli$group %in% c("ENG-SPA")], na.rm = TRUE))`, Range = `r printnum(min(stimuli$duration[stimuli$group %in% c("ENG-SPA")], na.rm = TRUE))`-`r printnum(max(stimuli$duration[stimuli$group %in% c("ENG-SPA")], na.rm = TRUE))`).

```{r stimulitable, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.cap="Stimuli data. Mean and standard deviation (SD) of lexical frequencies od target words (expressed in counts per million and as Zipf scores), density of higher-frequency phonological neighbourhoods of target words, and normalised Levenshtein phonological similarity between the presented and the target words."}

stimuli_table <- stimuli %>% 
  select(group, trial_id, frequency, frequency_zipf, pthn, lv) %>% 
  group_by(group) %>% 
  summarise_at(
    vars(frequency:lv), 
    list(
      mean = ~mean(., na.rm = TRUE),  
      sd = ~sd(., na.rm = TRUE)
    )
  ) %>%
  gt(rowname_col = "group") %>% 
  fmt_number(frequency_mean:lv_sd) %>% 
  tab_spanner(md("**Freq./million**"), c("frequency_mean", "frequency_sd")) %>% 
  tab_spanner(md("**Freq. (Zipf)**"), matches("frequency_zipf")) %>% 
  tab_spanner(md("***PTHN***"), matches("pthn")) %>% 
  tab_spanner(md("**Levenshtein**"), matches("lv")) %>% 
  cols_label(
    frequency_mean = md("Mean"),
    frequency_sd = md("*SD*"),
    frequency_zipf_mean = md("Mean"),
    frequency_zipf_sd = md("*SD*"),
    pthn_mean = md("Mean"),
    pthn_sd = md("*SD*"),
    lv_mean = md("Mean"),
    lv_sd = md("*SD*")
  ) %>% 
  as_latex()

stimuli_table

>>>>

author:
- name: Gonzalo Garcia-Castro
  affiliation: '1'
  corresponding: yes
  address: Ramon Trial Fargas, 25-27, 08005 Barcelona
  email: gonzalo.garciadecastro@upf.edu
- name: Serene Siow
  affiliation: '2'
- name: Nuria Sebastian-Galles
  affiliation: '1'
- name: Kim Plunkett
  affiliation: '2'
affiliation:
  - id            : "1"
    institution   : "Center for Brain and Cognition, Universitat Pompeu Fabra"
  - id            : "2"
    institution   : "Department of Experimental Psychology, University of Oxford"

authornote: |

abstract: |
  When listening to non-native speech, words whose translation in the native language is phonologically similar are easier to recognise. Previous studies suggest that low-proficiency bilinguals benefit more from this facilitation effect than high-proficiency bilinguals, but this claim has remained relatively unaddressed. In this study, we tested English and Spanish monolinguals in a translation elicitation task, in which they had to guess the translation of a list of words in an unfamiliar language (Spanish or Catalan). Presented words varied in their phonological similarity to their correct translation. This allowed us to test the impact of such phonological similarity on word recognition in a sample of listeners with very low proficiency in a non-native language. Translation accuracy increased substantially the likelihood of correct translations, but this effect was mediated by the size of the phonological neighbourhood of the correct translation. Participants benefitted from phonological similarity only when the correct translation was part of a sparse phonological neighbourhood. When the correct translation was part of a larger phonological neighbourhood, translation accuracy remained unaffected. Overall, these results provide evidence that litening to non-native words activated lexical representations in the native language, even in monolingual participants, resulting in competition for selection in the native language during translation.

  <!-- https://tinyurl.com/ybremelq -->

keywords: cognate, word recognition, translation, non-native, spoken word recognition, bayesian
wordcount: 4553

bibliography:
  - r-references.bib
  - references.bib
floatsintext: yes
figurelist: no
tablelist: no
footnotelist: no
linenumbers: yes
mask: no
draft: no
numbersections: yes
csl: "apa7.csl"
documentclass: apa7
classoption: man
output: papaja::apa6_docx
---

@@

set.seed(42)
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  results = "asis", 
  dpi = 1000
)

@@

Listening to non-native speech is more costly than listening to native speech, even for highly proficient bilinguals, and especially in acoustically adverse situations [@lecumberri2010non; @takata1990english]. One of the sources of this increased difficulty stems from the possible mismatch between the phonology of the native and the non-native languages: some acoustic features embedded in the non-native speech signal do not overlap perfectly with any phonemic category in the listener's native language. For example, a Spanish native listening to French may encounter the word /pɔʁt/ (*porte*), translation of *door* in French. The voiced uvular fricative consonant /ʁ/ and the open-mid back rounded vowel /ɔ/ do not exist in Spanish. These discrepancies can nonetheless be perceived as allophonic variations of native phonemes [@best2001discrimination], and the non-native speech signal engages lexical processing mechanisms [@weber2004lexical]. Although this mismatch has a noticeable toll on comprehension [@cutler2004patterns], the fact that word recognition can take place in such circumstances illustrates that non-native listeners--even low-proficiency ones--are rarely completely naïve to the language they are listening to. In the present study we investigate the extent to which adult listeners rely on the phonological similarity between the native and the non-native languages during word recognition, capitalising on the role of *cognates*.

All languages share, to some extent, similarities at the lexical level, frequently due to their typological closeness and/or socio-historical events involving the populations of speakers of each language. Cognates embody a particular case of such commonalities. Cognates are defined as cross-language synonyms whose form is similar (at the phonological level, orthographic level, signed level, etc.), frequently due to their shared etymological origin^[Some form-similar cross-language synonyms are technically not cognates. For example, *sun* and *sol* (in Spanish), share their phonological onset, but their etymology points to different origins. We will use the word *cognateness* to include all form-similar cross-language synonyms for simplicity. It is highly implausible that etymology plays a direct role on language perception if it is not via form-similarity, since it is not necessary for participants in psycholinguistic experiments to be aware of the etymology of the words they encounter in the tasks to be subject to the effect of form-similarity.]. For example, Romance languages such as Spanish and Catalan share many cognates [@schepens2012distributions], as in the case of *puerta* and *porta* (*door* in Spanish and Catalan, respectively). Cognates play a pivotal role in virtually all models of bilingual lexical processing because they provide evidence that bilinguals access the lexicon in a language non-selective way. For instance, pictures are named faster by Spanish-Catalan bilinguals when their associated labels in both languages are cognates (e.g., *puerta*-*porta*), compared to when their labels are non-cognates (*mesa*-*taula*, *table* in Spanish and Catalan), which suggests that word representations of both languages are activated during word retrieval [@costa2000cognate]. This cognate advantage is not restricted to production, but rather extends to word recognition [@thierry2007brain; @midgley2011effects], word learning [@de2000hard; @lotto1998effects; @valente2018does], and word translation [@christoffels2006memory].

Early theoretical accounts of this phenomenon pointed to the fact that the impact of cognateness on lexical processing depends on bilinguals' proficiency in their second language (L2) [@potter1984lexical; see @chen1989patterns for similar results]. The reason behind this claim was the assumption that second-language learners start acquiring words in L2 by associating them to their translation in L1 (lexical route), and not to their shared concept (conceptual route). This word-word connection would be sensitive to form-similarity (i.e., cognateness): the more similar the two word forms are, the stronger the connection). As learners become more proficient, the connection between L2 representations and their meaning grows stronger, and L2 word processing becomes less reliant on word-word connections between L1 and L2 representations, therefore to cognateness. The Revised Hierarchical Model [RHM, @kroll1994category] captured this hypothesis and predicted that translating words from L1 to L2 (*forward* translation) should take longer than translating from L2 to L1 (*backward* translation). The rationale behind this prediction was that backward translation relies more strongly on direct word-word links between L1 and L2 representations, whereas forward translation relies more strongly on the mediation between the concept and the two word forms. One of the consequences of this prediction is that backward translation should be more sensitive to the form similarity between the L1 and the L2 representations: cognate words should be retrieved faster than non-cognate words during backward translation.

To test these predictions, @degroot1994forward and @de1992determinants asked Dutch natives with high (but non-native) English proficiency to translate words from either Dutch to English (forward translation, L1 to L2) or from English to Dutch (backward translation, L2 to L1). Participants were presented with written words in English or Dutch, and were asked to speak out loud their translation in the other language as fast as possible. Results showed that translation times and accuracy were roughly equivalent in forward and backward translation. Cognates were translated equally fast in both conditions, but non-cognates were translated faster during backward than forward translation, and translation of cognates was less affected  by semantic variables than non-cognates. Moreover, forward translation was more sensitive than backward translation to the effect of semantic variables (e.g., concreteness). Overall, these findings supported a soft version of the RHM model, in which both the conceptual and the lexical translation routes are active during translation, but where backward translation relies more strongly on direct links between L1 and L2 representations, making it more sensitive to cognateness (than forward translation). Although higher-proficiency bilinguals tested in the same task showed similar results, subsequent studies did not find differences in participant's performance during forward and backward translation, or even found better performances in forward translation [@christoffels2006memory; @christoffels2013language]. These later findings put into question the RHM's predictions regarding the relationship between proficiency and cognateness, and backed a more complex interaction between the L1 and L2 lexica during word recognition.

Since the RHM model was proposed, later studies on monolingual populations brought attention to the role of the network of connections that words establish with each other at the form-level (phonological or orthographic) or at the conceptual level. For instance, the Neighbourhood Activation Model [NAM, @luce1998recognizing] highlighted the fact that lexical selection is affected by the number of phonological neighbours of the selected word. Words surrounded by a larger number of phonological neighbours (only differing in one phoneme from the target words) are responded more slowly and less accurately than words from sparser phonological neighbourhoods [@goldinger1989priming; @luce1990similarity], especially when such neighbours' lexical frequency is higher [@luce1998recognizing]. This finding posed an important question in the field of bilingualism research: do word representations in one language form part of the phonological or orthographic neighbours of the other language? The Bilingual Interactive Activation [BIA+, @van1998orthographic] addressed this issue and suggested that lexical representations from both languages establish both excitatory and inhibitory connections with other, resulting in an integrated lexicon for both languages, in line with connectionist approaches to lexical processing [@mcclelland1981interactive]. Evidence supporting such integrated lexicon is provided by the fact that bilinguals' performance in word recognition tasks is sensitive to the orthographic neighbourhood density of the translation of the presented word. When bilingual participants are presented with words in only one language (target language) in a lexical decision task, translating them to the non-target language took longer when they were part of large  neighbourhoods [@van1998orthographic]. This suggests that words presented in the target language activated orthographic neighbourhoods in the non-target language, which competed for selection with the target word during recognition.

A more recent model of bilingual lexical processing is Multilink [@dijkstra2019multilink], which integrates and formalises previous claims and predictions on how an interactive account of bilingual lexical access impacts word recognition, production, and translation. This model assumes an integrated lexicon in which word representations from the two languages establish connections in a way that is similar to what words of a single language do. Such model also assumes that during backward translation, the presented word in L2 activates form-similar words in L1, which in turn compete with the target word in L1. It also  proposes that translation equivalents are exclusively connected through their shared concept, contrary to the RHM model's specifications (which suggest that translation equivalents are connected by excitatory connections at the lexical level; e.g., translation can be done via word-word connections). In addition, this model suggests that the strength of the association between L2 representations and their concept is a function of language users' proficiency in L2. A remarkable correlation between Multilink simulations of a translation elicitation task and behavioural data on Dutch-English bilinguals [@christoffels2006memory] has been reported, showing that forward translation is faster than backward translation, contrary to the predictions made by the RHM model. Importantly, the model also generates data supporting the claim that cognates are translated faster than non-cognates, in line with previous literature on the cognate advantage during translation. Overall, results from Multilink are compatible with an integrated bilingual lexicon and with the existence of a facilitatory effect of cross-language similarity during translation.

The fact that Multilink successfully fitted the experimental data from @christoffels2006memory, despite the absence of any word-word connections between translation equivalents, is remarkable, and questions the necessity of such connections during translation, in favour of simpler models in which backward translation relies exclusively on the association between L2 and L1 word forms via their shared concept. It should be taken into account, however, that  @christoffels2006memory only tested fairly balanced bilinguals (i.e., bilinguals with high proficiency in L2), and that the model specification under which the quantitative predictions were generated, assumes similar vocabulary sizes across languages and same strength of word-concept connections of L1 and L2 representations. The aforementioned conclusions should therefore be taken with caution when extended to unbalanced bilinguals. It is low-proficiency bilinguals for whom the RHM model predicts strong reliance on word-word connections during translation, especially during backward translation, and this prediction remains untested my Multilink.

In the present study we investigated the plausibility of the lexical route as a mechanism exploited by monolingual adults during backward translation. Monolinguals can be considered a particular (extreme) case of unbalanced bilingualism, in which their vocabulary size in L2 is null, and word-concept connections are absent. In this case, monolinguals can only use the lexical route to succeed in the translation task, and they can only do so by exploiting the cognateness of the words presented and their translation in the native language. We hypothesised that if participants are able to translate words in L2, without knowledge of the language it belongs to, and based on their form-similarity with the target word, then the lexical route should be in place for low-proficiency bilinguals. In contrast, if participants are not able to translate words from L2 to L1 regardless of their form-similarity, then should low-proficiency bilinguals rely entirely on word-concept connections for backward translation.

Following @luce1998recognizing we also hypothesised that if participants are able to correctly translate words from an unfamiliar language based on their phonological similarity to the target word, similar sounding but incorrect translations (e.g., false friends) should be activated too. If this is the case, words whose translation in the native language are part of larger phonological neighbours should be less likely to be correctly translated, especially if such neighbours have higher lexical frequency than the target word.

@@

All materials, data, and code used in this study are hosted in the Open Science Framework [https://osf.io/9fjxm/](https://osf.io/9fjxm/) and a GitHub repository [https://github.com/bilingual-project/translation-elicitation.git](https://github.com/bilingual-project/translation-elicitation.git), along with additional notes.

@@

```{r stimulilengths, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
lengths <- stimuli %>% 
  select(group, ipa1, ipa2, word1, word2) %>% 
  mutate_at(vars(ipa1:word2), nchar) %>% 
  group_by(group) %>% 
  summarise_at(vars(ipa1, ipa2, word1, word2), list(mean = mean, sd = sd, min = min, max = max)) %>%
  split(.$group) %>% 
  map(select, -group) %>% 
  set_names(make_clean_names(names(.))) %>% 
  map(unlist)
```

@@

We retrieved PTHN scores (Phonological Total Higher-frequency Neighbourhood size) from the CLEARPOND database [@marian2012clearpond]. PTHN scores indicate the number of phonological neighbours of the target word with higher lexical frequency, as indicated by its score in the corresponding SUBTLEX database. CLEARPOND defines a phonological neighbour as a word whose phonological transcription in International Phonetic Alphabet (IPA) format (generated from eSPEAK, [http://espeak.sourceforge.net/](http://espeak.sourceforge.net/)) differs from that of the target word in only one addition, deletion, or substitution. PTHN scores in CLEARPOND measure have been calculated using corpora of similar size across language, allowing reliable cross-language comparisons.

@@

```{r stimuli-correlations, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.width=6, fig.height=4, fig.cap="Marginal distribution and pairwise correlation of lexical frequency, PTHN, and Levenshtein similarity. Statistics are indicated for the three experimental conditions (ENG-SPA, ENG-CAT, and SPA-CAT) with different colours. The upper triangle of the plot matrix shows the Pearson correlation between each pair of variables (asterisks indicate statistical significance at .05, .01, and .001 significance thresholds). The diagonal of the plot matrix indicates the marginal distribution of eacch variable. The lower triangle of the plot matrix representes graphically the association between each pair of variables, together with a linear regression line."}
my_bin <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_density(..., size = 1, alpha = 0.25) +
    geom_rug(...)
}

my_smooth <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(..., shape = 1, stroke = 1, alpha = 0.25, size = 0.5) +
    geom_smooth(..., method = "lm", formula = y ~ x, alpha = 0.25, size = 1)
}

my_cor <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    ggally_cor(..., size = 7, title = "r", )
}

stimuli %>% 
  drop_na(frequency_zipf, pthn, lv) %>% 
  ggpairs(
    mapping = aes(colour = group, fill = group),
    columns = c("frequency_zipf", "pthn", "lv"),
    columnLabels = c("Frequency (Zipf)", "PTHN", "Levenshtein"),
    lower = list(continuous = my_smooth),
    diag = list(continuous = my_bin),
    upper = list("my_cor")
  ) +
  scale_fill_manual(values = c("#1A85FF", "#ff2976", "#FFC20A"), na.translate = FALSE) +
  scale_color_manual(values = c("#1A85FF", "#ff2976", "#FFC20A"), na.translate = FALSE)

Participants listened to one audio file in each trial. This audio file corresponded to a word in Catalan (for Spanish speakers, and for English participants allocated in the Catalan condition) or Spanish (for English speakers allocated in the Spanish condition). The audio files were the same ones used in child experiments conducted in the Laboratori de Recerca en Infància of Universitat Pompeu Fabra (Barcelona, Spain). These audio files were recorded by a proficient Catalan-Spanish female bilingual from the Metropolitan Area of Barcelona in a child-directed manner. Catalan and Spanish words were recorded at 44,100 Hz in separate files in the same session, and then de-noised using Audacity and normalised at peak intensity using Praat [@broersma2021praat]. The average duration of the audios was `r printnum(mean(stimuli$duration, na.rm = TRUE))` (*SD* = `r printnum(sd(stimuli$duration, na.rm = TRUE))`, Range = `r printnum(min(stimuli$duration, na.rm = TRUE))`-`r printnum(max(stimuli$duration, na.rm = TRUE))`). The average duration of the Catalan audio files was `r printnum(mean(stimuli$duration[stimuli$group %in% c("ENG-CAT")], na.rm = TRUE))` seconds (*SD* = `r printnum(sd(stimuli$duration[stimuli$group %in% c("ENG-CAT")], na.rm = TRUE))`, Range = `r printnum(min(stimuli$duration[stimuli$group %in% c("ENG-CAT")], na.rm = TRUE))`-`r printnum(max(stimuli$duration[stimuli$group %in% c("ENG-CAT")], na.rm = TRUE))`), and the average duration of the Spanish audios was `r printnum(mean(stimuli$duration[stimuli$group %in% c("ENG-SPA")], na.rm = TRUE))` seconds (*SD* = `r printnum(sd(stimuli$duration[stimuli$group %in% c("ENG-SPA")], na.rm = TRUE))`, Range = `r printnum(min(stimuli$duration[stimuli$group %in% c("ENG-SPA")], na.rm = TRUE))`-`r printnum(max(stimuli$duration[stimuli$group %in% c("ENG-SPA")], na.rm = TRUE))`).

```{r stimulitable, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
